\bf Longitudinal Principal Manifold Estimation

### 2024-07-10 paper/main.tex
overleaf updates

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url}

%\documentclass[12pt]{amsart}
%\documentclass[11pt,reqno]{article}
%\usepackage{cases}

%\RequirePackage[numbers]{natbib}
%\RequirePackage[authoryear]{natbib}%% uncomment this for author-year citations
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}%% uncomment this for coloring bibliography citations and linked URLs
%\RequirePackage{graphicx}%% uncomment this for including figures

%\usepackage[
%backend=biber,
%style=authoryear,
%url=false,
%isbn=false,
%date=year,
%doi=false
%]{biblatex}

%\addbibresource{references.bib}

\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amscd}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{mathrsfs}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage[perpage,symbol*]{footmisc}
\usepackage{float}
\usepackage{hyperref}
\usepackage{color}
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{bm}

\usepackage{algorithm2e}

%\usepackage{natbib}
%\bibliographystyle{abbrvnat}
%\setcitestyle{authoryear,open={(},close={)}}
\usepackage{authblk}

\usepackage{csquotes}
\usepackage[english]{babel}

\renewcommand{\baselinestretch}{1.0}
\setlength{\oddsidemargin}{-0.5cm}
\setlength{\evensidemargin}{-0.5cm}
\renewcommand{\topmargin}{-2cm}
\renewcommand{\oddsidemargin}{0mm}
\renewcommand{\evensidemargin}{0mm}
\renewcommand{\textwidth}{180mm}
\renewcommand{\textheight}{240mm}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\T}{\intercal}
\newcommand{\commentout}[1]{}

\newcommand{\kmedit}[1]{{\color{purple}  #1}}
\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}

\newcommand{\zielinski}[1]{{\color{blue} \sf $\spadesuit\spadesuit\spadesuit$ Rob Zielinski: [#1]}}

@@

@@

\title{Longitudinal Principal Manifold Estimation}
\author[1]{Robert Zielinski}
\author[2]{Kun Meng}
\author[1]{Ani Eloyan}
\author[*]{for the Alzheimer’s Disease Neuroimaging Initiative}
\affil[1]{Department of Biostatistics, Brown University}
\affil[2]{Division of Applied Mathematics, Brown University}

\maketitle

\doublespacing

\section*{Abstract}

Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation. \footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}

% Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This approach mitigates the effects of random disturbances to the manifold between time points. In addition, we propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset. We present a comprehensive analysis of magnetic resonance imaging data from 263 participants demonstrating high levels of variability of regional volumes when using existing methods for segmentation of hippocampuses and thalamnuses and the advantages gained by using our proposed methods that leverage longitudinal regional trends for segmentation.  \footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}

\section{Introduction}

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) and Parkinson's disease (\cite{knopmanAlzheimerDisease2021}, \cite{poeweParkinsonDisease2017}). Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data (\cite{crainiceanu2016tutorial}). For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes. Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability (\cite{boccardiSurveyProtocolsManual2011}). Automated segmentation approaches, such as FSL-FIRST and FreeSurfer, have been introduced to address these concerns (\cite{patenaudeBayesianModelShape2011}, \cite{reuterWithinsubjectTemplateEstimation2012}). While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

\subsection{A Motivating Question from Medical Imaging}

Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models. In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.

Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

\subsection{A Brief Overview of Existing Approaches and Our Contributions}

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}). Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}), single-cell biology (\cite{ding2023learning}), and robotics (\cite{gao2023k}; \cite{gao2024bi}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.

Nonlinear manifold learning methods have been extensively studied in the literature. However, many of them are not applicable in longitudinal studies, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}). That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of the time dimension. We use the oldest linear manifold learning method, principal component analysis (PCA, \cite{Pearson1901on}), as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. In addition, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation~\eqref{eq: spline representation of the PME}, in Section \ref{s:PME}). Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

@@

This work was motivated by longitudinal imaging studies, where data of participants are collected during several study visits to observe changes in brain structure (among other variables of interest) over time. One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this article. The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}. The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources, including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers. We implement the proposed methods to estimate the surfaces of two regions of interest in the brain - the hippocampus and the thalamus. The thalamus is selected as a ``easy-to-fit" region of interest to evaluate the performance of our proposed methods when the underlying structure is somewhat close to a spherical shape. The hippocampus is more of a pear-shaped region and is selected as a region that may exhibit atrophy for people with AD.

The remainder of this article is laid out as follows. In Section \ref{s:PME}, we review the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}. In Section \ref{s:LPME}, we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show our proposed novel  algorithm for dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in individuals with AD and cognitively normal adults. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem making this a supervised learning problem. \cite{mengPrincipalManifoldEstimation2021} also show that this function takes spline form
\begin{align}\label{eq: spline representation of the PME}
    f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ \ \text{ for } \ l = 1, 2, \dots, D,
\end{align}
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud. Importantly, $N$ tends to be much smaller than the number of points in the data cloud (see Figure 3 of \cite{mengPrincipalManifoldEstimation2021}), which indicates that the representation in equation~\eqref{eq: spline representation of the PME} is computationally efficient. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

\section{Longitudinal Principal Manifold Estimation Algorithm}\label{s:LPME}

@@

\newpage

\appendix

\section*{Appendix}

>>>>

\documentclass[12pt]{article}
\usepackage{algorithm2e}
\usepackage{amscd}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{bm}
\usepackage{caption}
\usepackage{color}
%\usepackage{csquotes}
\usepackage{float}
\usepackage[perpage, symbol*]{footmisc}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{url}

\newcommand{\blind}{1}
\graphicspath{{figures/}}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\T}{\intercal}

@@

\setcitestyle{aysep={}}

@@

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Longitudinal Principal Manifold Estimation}
  \author{
    Robert Zielinski\hspace{.2cm}\\
    Department of Biostatistics, Brown University\\
    Kun Meng \\
    Division of Applied Mathematics, Brown University\\
    and \\
    Ani Eloyan \\
    Department of Biostatistics, Brown University\\
    for the Alzheimer's Disease Neuroimaging Initiative}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Longitudinal Principal Manifold Estimation}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas susceptible to atrophy in those with neurodegenerative conditions like Alzheimer’s disease. Most methods for extracting brain regions are applied to scans from study participants independently, resulting in wide variability in shape and volume estimates of these regions over time in longitudinal studies. To address this problem, we propose a longitudinal principal manifold estimation method, which seeks to recover smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of principal manifold embedding functions estimated at each time point. This mitigates the effects of random disturbances to the manifold between time points. Additionally, we propose a novel data augmentation approach to enable principal manifold estimation on self-intersecting manifolds. Simulation studies demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. The proposed method improves the estimation of surfaces of hippocampuses and thalamuses using data from participants of the Alzheimer’s Disease Neuroimaging Initiative. An analysis of magnetic resonance imaging data from 236 individuals shows the advantages of our proposed methods that leverage regional longitudinal trends for segmentation. %\footnote{\textbf{Abbreviations:} AD, Alzheimer's disease; ADNI, Alzheimer's Disease Neuroimaging Initiative; HDMDE, high-dimensional mixture density estimation; LPME, longitudinal principal manifold estimation; MRI, magnetic resonance imaging; PCA, principal component analysis; PME, principal manifold estimation;}
\end{abstract}

\noindent%
{\it Keywords:}  Alzheimer's disease, smoothing spline, magnetic resonance imaging
\vfill

\newpage
\spacingset{1.9}
\section{Introduction}

%Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}. Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}. For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes. Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Additionally, there are a large number of manual segmentation protocols, each producing distinct results and incurring varying levels of inter-rater and intra-rater variability \citep{boccardiSurveyProtocolsManual2011}. Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}. While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

Neuroimaging plays a critical role in the diagnosis and monitoring of a number of common neurodegenerative conditions, such as Alzheimer's disease (AD) \citep{knopmanAlzheimerDisease2021}. Frequently, interest centers around longitudinal changes in one or more neurological substructures. These structural changes can be observed in magnetic resonance imaging (MRI) data \citep{crainiceanu2016tutorial}. For example, it is common to observe atrophy in the hippocampus, a structure in the temporal lobe of the brain, of those with  cognitive impairment either due to AD or other causes. Image segmentation enables the extraction of subcortical structures from MRI images of the brain for identifying differences between these structures in disease populations, modeling trajectories of change in the structures, and evaluating treatment effects in terms of reduction of atrophy over time. Traditionally, manual segmentation of images by a trained radiologist has been considered the most accurate approach for segmentation of regions of interest and is the gold standard. However, this approach is highly time and resource intensive. When analyzing data from studies with a large number of images, this approach may impose prohibitive costs. Automated segmentation approaches, including FSL-FIRST, have been introduced to address these concerns \citep{patenaudeBayesianModelShape2011}. While automating the segmentation process drastically reduces costs, it potentially introduces additional inaccuracies.

%\subsection{A Motivating Question from Medical Imaging}

%Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. %\cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with AD for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the brain images of cognitively normal people which compose the majority of the data used for training the segmentation models. %In studying longitudinal trajectories of atrophy, segmentation problems may result in seemingly random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging \citep{scahill2003longitudinal}. %To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice. A thorough description of the data and computation of hippocampus volumes and further examples are given in Section \ref{s:application} and Figure \ref{fig:lhipp_volume_comparison}.

Failures of automatic segmentation algorithms may have dramatic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a study comparing the accuracy of FIRST and FreeSurfer, another automated segmentation method, using a large sample, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing the scans with failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. Ultimately, high levels of variability are present at the subcortical level, both between study visits for the same individual and between individuals, regardless of the segmentation approach, and may be particularly influential when using automated segmentation methods. Mitigating the extent of this variability is a priority from a statistical perspective to improve the inferences that can be made from segmented image data. To this end, in this article, we propose a manifold learning-based method to develop smooth estimates of the surfaces of subcortical structures over time.

%\subsection{Background and Our Contributions}

%Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing \citep{shen2022robust}, single-cell biology \citep{ding2023learning}, and robotics \citep{gao2024bi}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} \citep{fefferman2016testing}. Manifold learning approaches have been applied in various scientific fields, e.g., single-cell biology \citep{ding2023learning} and robotics \citep{gao2023k}. In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures. Nonlinear manifold learning methods have been extensively studied in the literature. However, many popular approaches, such as Isomap \citep{tenenbaumGlobalGeometricFramework2000}, locally linear embedding \citep{roweisNonlinearDimensionalityReduction2000}, and Laplacian eigenmaps \citep{belkin2003laplacian}, are not applicable in longitudinal studies. That is, these methods include the time dimension in the dimension reduction process, causing the time dimension to be incorporated into the low-dimensional parameterization and preventing meaningful interpretation of the time dimension. %We use the oldest linear manifold learning method, principal component analysis \citep[PCA,][]{Pearson1901on}, as a simple example to illustrate this point: A principal component may be a linear combination of both time and spatial coordinates.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal PCA method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. As for nonlinear methods, \cite{wolzManifoldLearningBiomarker2010} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space. However, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. Additionally, \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

%In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}). Importantly, our approach allows us to iteratively update the smoothing and fitting steps, which could lead to an overall better fit. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

In this article, we propose a new longitudinal approach to estimating nonlinear manifolds over time. Rather than treating the time dimension the same as the space dimensions, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed by \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most popular manifold learning methods, this approach results in an analytic and computationally efficient functional representation of the manifold estimate (see equation \eqref{eq: spline representation of the PME}). We build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between time points. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, and 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

@@

This work was motivated by longitudinal imaging studies, where data of participants are collected during several study visits to observe changes in brain structure (among other variables of interest) over time. One such study is the ADNI, where we obtained data that we analyzed to reach the results presented in this article. The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}. The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources, including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers. We implement the proposed methods to estimate the surfaces of two regions of interest in the brain - the hippocampus and the thalamus. The thalamus is selected as a ``easy-to-fit" region of interest to evaluate the performance of our proposed methods when the underlying structure is somewhat close to a spherical shape. The hippocampus is more of a pear-shaped region and is selected as a region that may exhibit atrophy for people with AD.

The remainder of this article is laid out as follows. Section \ref{s:PME} reviews the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}. In Section \ref{s:LPME}, we adapt the principal manifold framework for longitudinal settings, and show our proposed novel algorithm for dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in ADNI participants. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation \eqref{eq:pme_kappa} consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, even though manifold learning is commonly considered an unsupervised learning approach, \cite{mengPrincipalManifoldEstimation2021} show that finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in fact takes the form of a penalized regression problem, making this a supervised learning problem. \cite{mengPrincipalManifoldEstimation2021} also show that this function takes spline form
\begin{align}\label{eq: spline representation of the PME}
    f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ \ \text{ for } \ l = 1, 2, \dots, D,
\end{align}
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud. Importantly, $N$ tends to be much smaller than the number of points in the data cloud \citep[fig. 3]{mengPrincipalManifoldEstimation2021}, which indicates that the representation in equation \eqref{eq: spline representation of the PME} is computationally efficient. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

\section{Longitudinal Principal Manifold Estimation}\label{s:LPME}

@@

%\if1\blind
%{
%\section*{Acknowledgment}
